Automatic Identification and Annotation of Tasks in Structured Programs

### 2018-02-17 docs/ACM/paper.tex
Adding example to section 3.4.

\begin{example}
\label{ex:cond_task}
The condition in line 7 of Fig.~\ref{fig:ex_Regions} determines that tasks can be
created if \textsf{5 $\times$ M $<$ WORK\_CUTOFF}
holds.
The expression \textsf{5 $\times$ M} is an approximation for the size of the
task that comprises the loop at lines 8-10.
The constant \textsf{WORK\_CUTOFF} represents the cost of creating a thread
in the OpenMP runtime, and it is determined empirically.
\end{example}

\begin{figure}[t!]
\begin{small}
\begin{eqnarray*}
\begin{tabular}{lc}
\textsc{[Instr]} &
W(\mbox{instr}) = \mbox{given by the spec.}
\\\\
\textsc{[Seq]} &
\inferrule{W(S_1) = w_1 \\ W(S_2) = w_2}{W(S_1;S_2;) = w_2}
\\\\
\textsc{[Branch]} &
\inferrule{W(S_1) = w_1 \\ W(S_2) = w_2 \\ W(S_3) = w_3}
{W(\mbox{if}(S_1) \ S_2; \ \mbox{else} \ S_3;) = w_1 + w_2 + w_3}
\\\\
\textsc{[LoopInf]} &
\inferrule{W(S_1) = w_1 \\ W(S_2) = w_2 \\ X(S_1) = \infty}
{W(\mbox{while} (S_1) \ S_2;) = 10 \times (w_1 + w_2)}
\\\\
\textsc{[LoopExp]} &
\inferrule{W(S_1) = w_1 \\ W(S_2) = w_2 \\ X(S_1) = E}
{W(\mbox{while} (S_1) \ S_2;) = E \times (w_1 + w_2)}
\end{tabular}
\end{eqnarray*}
\end{small}
\caption{\label{fig:swe}Estimating the cost of tasks.}
\end{figure}

% Explain how we use symbolic information to compute costs.
The declarative rules in Figure~\ref{fig:swe} sketches the heuristics that
we use to compute the SWE for a given hammock region $S$.
The only rules that deserves comments are \textsc{LoopInf} and
\textsc{LoopExp}.
The former applies on loops whose number of iterations we cannot bound via some
symbolic expression computed statically.
The latter represents loops that we can analyze using our Symbolic Range Analysis.
The auxiliary function $X(S)$ returns a symbolic expression that represents
the range of values covered by the code $S$ that controls the number of
iterations of the loop.
In Figure~\ref{fig:alg_main}, these rules are implemented by function
\textsf{profitabilityAnalysis}.

@@

Task expansion consists in finding the smallest task that is large enough to
pay for the thread creation cost.
Figure \ref{fig:expand_alg} shows the algorithm that performs this activity.
We start this algorithm by setting \textsf{REGION} to be a hammock graph $H$
that is contained entirely within a windmill $W$.
The hammock decomposition of a structured program forms a tree~\cite{Ferrante87},
such that vertices in this tree represent hammock graphs.
$H_1$ is a child of $H_2$ in this tree if $H_1 \subset H_2$, and for any other
node $H_x$, $H_1 \subset H_x \Rightarrow H_2 \subset H_x$.
In this context, we call $H_2$ the \textsf{PARENT} of $H_1$.

\begin{definition}[Region expansion]
\label{def:expansion}
A task region $R$ can be expanded into a larger region $R'$ if, and only if:
\begin{itemize}
\item $|M(R')| \leq |M(R)|$, in which $M(R)$ and $M(R')$ are the set of input and output memory
dependences in regions $R$ and $R'$ respectively.
\item $R'$ is still inside a loop or a cyclic construct.
\end{itemize}
\end{definition}

Figure \ref{fig:expand_alg} illustrate the algorithm of expansion. It simply tries to expand while
all the conditions in Definition \ref{def:expansion} are met, and then halts
when they are not. After this process, {\Taskminer} ends up with a list
of potential \emph{task regions}, which will then pass under
a series of operations prior to the annotation to source code, the 
ultimate step.

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\columnwidth]{images/expand_alg}
\caption{Task discovery via expansion of hammock regions.}
\label{fig:expand_alg}
\end{center}
\end{figure}

>>>>

\begin{example}
\label{ex:cond_task}
The condition in line 7 of Fig.~\ref{fig:ex_Regions} determines that tasks are
created if \textsf{5 $\times$ M $<$ WORK\_CUTOFF}.
\textsf{M} is a program symbol, i.e., a variable passed as argument to
function \textsf{foo}.
The expression \textsf{5 $\times$ M} is an approximation for the size of the
task that comprises the loop at lines 8-10.
The constant \textsf{WORK\_CUTOFF} represents the cost of creating a thread
in the OpenMP runtime, and it is determined empirically.
\end{example}

\begin{figure}[t!]
\begin{small}
\begin{eqnarray*}
\begin{tabular}{lc}
\textsc{[Instr]} &
W(\mbox{instr}) = \mbox{given by the spec.}
\\\\
\textsc{[Seq]} &
\inferrule{W(S_1) = w_1 \\ W(S_2) = w_2}{W(S_1;S_2;) = w_2}
\\\\
\textsc{[Branch]} &
\inferrule{W(S_1) = w_1 \\ W(S_2) = w_2 \\ W(S_3) = w_3}
{W(\mbox{if}(S_1) \ S_2; \ \mbox{else} \ S_3;) = w_1 + w_2 + w_3}
\\\\
\textsc{[LoopInf]} &
\inferrule{W(S_1) = w_1 \\ W(S_2) = w_2 \\ \mathit{Iter}(S_1) = \infty}
{W(\mbox{while} (S_1) \ S_2;) = 10 \times (w_1 + w_2)}
\\\\
\textsc{[LoopExp]} &
\inferrule{W(S_1) = w_1 \\ W(S_2) = w_2 \\ \mathit{Iter}(S_1) = E}
{W(\mbox{while} (S_1) \ S_2;) = E \times (w_1 + w_2)}
\end{tabular}
\end{eqnarray*}
\end{small}
\caption{\label{fig:swe}Estimating the cost of tasks.}
\end{figure}

% Explain how we use symbolic information to compute costs.
The declarative rules in Figure~\ref{fig:swe} sketches the heuristics that
we use to compute the SWE for a given hammock region $S$.
Rule \textsc{LoopInf} applies on loops whose number of iterations we cannot bound
via some symbolic expression computed statically.
Rule \textsc{LoopExp} represents loops that we can analyze using our Symbolic
Range Analysis.
The auxiliary function $\mathit{Iter}(S)$ returns a symbolic expression that
represents
the range of values covered by the code $S$ that controls the number of
iterations of the loop.
In Figure~\ref{fig:alg_main}, these rules are implemented by function
\textsf{profitabilityAnalysis}.

@@

Task expansion consists in finding the smallest task that is large enough to
pay for the thread creation cost.
Figure \ref{fig:expand_alg} shows the algorithm that performs this activity.
We start this algorithm by setting \textsf{REGION} to be a hammock graph $H$
that corresponds to a vane within a windmill $W$.
The hammock decomposition of a structured program forms a tree~\cite{Ferrante87},
such that vertices in this tree represent hammock graphs.
$H_1$ is a child of $H_2$ in this tree if two conditions apply:
(i) $H_1 \subset H_2$, and (ii) for any 
other node $H_x$, if $H_1 \subset H_x$ then $H_2 \subset H_x$.
In this context, we call $H_2$ the {\em parent} of $H_1$.

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\columnwidth]{images/expand_alg}
\caption{Task discovery via expansion of hammock regions.
\textsf{COST} is the overhead of creating and scheduling threads, as determined
by a cost model.}
\label{fig:expand_alg}
\end{center}
\end{figure}

The boolean \textsf{CanExpand} in Figure~\ref{fig:expand_alg} is determined by
the routine \textsf{checkExpansion}.
This function remains true for as long as the following conditions apply with
respect to its input \textsf{VANE}:
(i) \textsf{VANE} is a hammock region with the structural properties of a
windmill's vane, i.e., it is contained within a windmill $W$;
(ii) we can analyze all the memory regions within \textsf{VANE} using the
symbolic range analysis of Section~\ref{sub:sra};
(iii) \textsf{VANE} does not depend on another vane inside the same windmill $W$.
In the rest of this section, we shall use Figure~\ref{fig:ex_expansion} to
explain these conditions.

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_expansion}
\caption{(a) Example of doubly nested loop that shall be analyzed by the 
\Taskminer.
(b) The dependence graph of the program.
(c) The decomposition of this loop into windmills and vanes.
$D$ is the set of variables the vane depends on.}
\label{fig:ex_expansion}
\end{center}
\end{figure}

The program in Figure~\ref{fig:ex_expansion} (a), which is a slight variation of
the code earlier seen in Figure~\ref{fig:ex_regions}, contains two windmills.
The first consists in the \textsf{for} loop at line 1.
The second, nested in the first, is the \textsf{for} loop at line 2.
The vanes that sprout from these windmills, are highlighted in the program
dependence graph seen in Figure~\ref{fig:ex_expansion} (b).
The figure shows that we have three task candidates, each one formed by a
different vane.
In order to find the actual tasks, we apply function \textsf{expand}
(Fig.~\ref{fig:expand_alg}) onto the innermost vanes: \textsf{Vane-2} and
\textsf{Vane-3}.
The latter cannot be further expanded, because it depends on the former.
On the other hand, \textsf{Vane-2} can be expanded to include also
\textsf{Vane-3}, as this expansion does not create new dependences between vanes
in the same windmill.
